---
title: "Power_analysis"
format:
  gfm:
    toc: true
    number-sections: true
    colorlinks: true
editor: visual
---

## *Power Analysis*

*A simulation-based power analysis was conducted for a Bayesian multilevel mediation model fit with `brms` to evaluate the sample size required to detect direct and indirect effects in a 14-day intensive longitudinal design. The model included two daily predictors, **stress** and **importance of the intention**, two daily mediators, **offloading** and **monitoring**, and a binary end-of-day outcome, **prospective memory**.*

*The data-generating model assumed Bernoulli outcomes and incorporated both fixed and random effects. Predictor-to-mediator paths were set to moderate effects on the log-odds scale (`a1 = -0.40`, `a2 = 0.40`, `b1 = -0.40`, `b2 = 0.40`). Mediator-to-outcome paths were also set to moderate effects (`d1 = 0.40`, `d2 = 0.40`), whereas direct predictor-to-outcome effects were set to smaller values (`c1 = -0.20`, `c2 = 0.20`). \
This setting produced indirect effects of approximately `0.16` for each mediated pathway (i.e., `0.40 × 0.40 = 0.16`), corresponding to small-to-moderate indirect effects on the log-odds scale.*

*The simulation further included person-level random intercepts for offloading, monitoring, and prospective memory (`SD = 1.0`), as well as modest random slopes (`SD = 0.10`) for the effects of offloading, monitoring, and stress on prospective memory, and for the effects of stress and importance on offloading and monitoring. This structure was chosen to allow between-person variability while maintaining an estimable multilevel model.*

*The entire code used to generate is showed here:*

```{r}
#| eval: true
#| message: false
#| warning: false
#| include: true
# ---- Packages ----
library(MASS)    # mvrnorm
library(dplyr)
library(tidyr)
library(brms)

# ---- 1) Data simulator for 14-day intensive longigudinatl study ----

# the three dependent vars, PM, offloading, and monitoring are binary. 
# The approach that I am using is considering them as continuous first, 
# and then binarize them depending on their probability of occurring (marginal prevalence) -
simulate_ild_pm <- function( 
                            T_per = 14, # days
                            seed = 123,
                            # fixed effects (within-person daily)
                            intercept_pm = 0, # intercept pm (log odds) = p = 0.24 to convert, exp(log-odds)/(1(1+exp(log-odds)). this is chance  p =0.5 
                            # to convert probability to logit: log(p/(1-p))
                            
                            # coefficients
                            a1 = - 0.40,   # stress -> offloading
                            a2 = 0.40,   # importance -> offloading
                            b1 = - 0.40,   # stress -> monitoring
                            b2 = 0.40,   # importance -> monitoring
                            c1 = - 0.20,   # direct stress -> pm (log odds - medium = 0,.5)
                            c2 = 0.20,   # direct importance -> pm (log odds)
                            d1 = 0.40,  # offloading -> pm 
                            d2 = 0.40,   # monitoring -> pm
                            
                            # person-level random effect SDs a
                            re_sds = c(off = 1.0, mon = 1.0, pm = 1.0, # random intercepts
                                       pm_off_slope = 0.1, pm_mon_slope = 0.1, 
                                       pm_stress_slope = 0.1, a1_slope = 0.1, b1_slope = 0.1, 
                                       a2_slope = 0.1, b2_slope = 0.1),
                            
                            # covariance structure between person-level random effects (4x4)
                            re_corr_mat = NULL,
                            
                            # whether to include a random slope for offloading -> pm
                            include_pm_off_slope = TRUE,
                            
                            # AR structure for stress/importance (within-person temporal autocorr)
                            ar_stress = 0.35,
                            ar_importance = 0.35,
                            miss_prob = 0.0) {
  
  set.seed(seed)
  TT <- T_per
  
  # ---- build list of random effects 
  re_names_expected <- c("u_off","u_mon","u_pm", # random intercepts
                         "u_pm_off_slope","u_pm_mon_slope","u_pm_stress_slope", # slopes
                         "u_a1_slope","u_b1_slope","u_a2_slope","u_b2_slope") # slopes
  
  # ensure re_sds named; fill missing with 0 and warn
  if (is.null(names(re_sds))) names(re_sds) <- rep("", length(re_sds))
  
  # Map missing expected names to 0
  for (nm in re_names_expected) if (!(nm %in% names(re_sds))) {
    # try to accept shorter keys (e.g., "off" -> u_off)
    alt <- sub("^u_", "", nm)
    if (alt %in% names(re_sds)) {
      names(re_sds)[names(re_sds) == alt] <- nm
    } else {
      re_sds[[nm]] <- 0
      warning(sprintf("re_sds did not contain '%s' — defaulting SD=0 for this RE.", nm))
    }
  }
  
  # now construct sds_vec in the expected order
  sds_vec <- as.numeric(re_sds[re_names_expected])
  names(sds_vec) <- re_names_expected
  
  kRE <- length(sds_vec)
  
  # correlation matrix
  re_corr_mat <- diag(kRE)
  
  # create covariance matrix from sds and correlation matrix
  RE_cov <- diag(sds_vec) %*% re_corr_mat %*% diag(sds_vec)
  
  # eign value
  ev <- eigen(RE_cov, symmetric = TRUE)$values
  
  # intercept handling
  if (is.null(intercept_pm)) intercept_pm <- qlogis(intercept_prob)
  
  # Draw person-level random effects jointly: each person has vector (u_off, u_mon, u_pm, u_pm_off_slope)
  REs_full <- MASS::mvrnorm(N, mu = rep(0, length(sds_vec)), Sigma = RE_cov)                                                  
  
  colnames(REs_full) <- re_names_expected
  
  # storage
  dat_list <- vector("list", N)
  
  
  for (i in seq_len(N)) { # loop through particiapnts
    # simulate time series for exogenous predictors (stress & importance) AR(1)
    stress <- numeric(TT)
    importance <- numeric(TT)
    
    for (t in seq_len(TT)) {
      if (t == 1) {
        stress[t] <- rnorm(1, 0, 1)
        importance[t] <- rnorm(1, 0, 1)
      } else {
        stress[t] <- ar_stress * stress[t-1] + rnorm(1, 0, sqrt(1 - ar_stress^2))
        importance[t] <- ar_importance * importance[t-1] + rnorm(1, 0, sqrt(1 - ar_importance^2))
      }
    }
    
    # person-level REs
    # extract REs for person i (0 if SD==0)
    u_off <- REs_full[i,"u_off"]; u_mon <- REs_full[i,"u_mon"]; u_pm <- REs_full[i,"u_pm"]
    u_pm_off_slope <- REs_full[i,"u_pm_off_slope"]; u_pm_mon_slope <- REs_full[i,"u_pm_mon_slope"]
    u_pm_stress_slope <- REs_full[i,"u_pm_stress_slope"]
    u_a1_slope <- REs_full[i,"u_a1_slope"]; u_b1_slope <- REs_full[i,"u_b1_slope"]
    u_a2_slope <- REs_full[i,"u_a2_slope"]; u_b2_slope <- REs_full[i,"u_b2_slope"]
    
    off_v <- integer(T_per); mon_v <- integer(T_per); pm_v <- integer(T_per)
    
    
    for (t in seq_len(TT)) { # loop through the trials
      
      # Mediator linear predictors (logit scale) include person intercept
      eta_off <- u_off + (a1 + u_a1_slope) * stress[t] + (a2 + u_a2_slope) * importance[t]
      p_off <- plogis(eta_off)
      off_t <- rbinom(1, 1, p_off)
      
      eta_mon <- u_mon + (b1 + u_b1_slope) * stress[t] + (b2 + u_b2_slope) * importance[t]
      p_mon <- plogis(eta_mon)
      mon_t <- rbinom(1, 1, p_mon)
      
      # PM linear predictor uses observed mediators (binary) and person REs
      # include random slope u_pm_off_slope if requested
      # PM eqn with person-specific slopes for stress and both mediators
      slope_off_term <- d1 + u_pm_off_slope
      slope_mon_term <- d2 + u_pm_mon_slope
      stress_term    <- c1 + u_pm_stress_slope
      
      eta_pm <- intercept_pm + u_pm + stress_term * stress[t] + c2 * importance[t] +
        slope_off_term * off_t + slope_mon_term * mon_t
      p_pm <- plogis(eta_pm)
      pm_t <- rbinom(1, 1, p_pm)
      
      
      # store
      off_v[t] <- off_t
      mon_v[t] <- mon_t
      pm_v[t]  <- pm_t
    }
    
    dat_list[[i]] <- data.frame(id = i, time = 1:T_per,
                                stress = stress, importance = importance,
                                offloading = off_v, monitoring = mon_v,
                                prospective_memory = pm_v)
  }
  
  dat <- bind_rows(dat_list)
  
  # optionally drop person-level RE columns for analysis (keeps only observed variables)
  dat <- dat %>% select(-starts_with("u_"))
  
  if (miss_prob > 0) {
    set.seed(seed + 1)
    dat <- dat[runif(nrow(dat)) > miss_prob, ]
  }
  
  return(dat)
}


```

*Power was evaluated across a range of sample sizes (from N=20 to N=220), by repeatedly simulating datasets, fitting the model, and calculating the proportion of replications in which each target parameter was detected. On each iteration, a target parameter was considered detected if the posterior probability of the effect being in the expected direction exceeded 0.95. Specifically, for positive directional hypotheses, detection was defined as the proportion of posterior draws greater than zero exceeding 0.95, and for negative directional hypotheses, detection was defined as the proportion of posterior draws less than zero exceeding 0.95.*

A code running the simulations and calculating the power is the following:

```{r}
#| eval: false
#| include: false
#| echo: true
#| message: false
#| warning: false

set.seed(100)
dat_example <- simulate_ild_pm(N = 80, T_per = 14, seed = 100)

table(dat_example$offloading)  # marginal counts
table(dat_example$monitoring)
table(dat_example$prospective_memory)

# ---- brms model specification consistent with the simulator ----

# allow brms to estimate correlated person-level random intercepts
# (and random slope for offloading in PM).
bf_off  <- bf(offloading ~ stress + importance + (1+stress+importance | id), 
              family = bernoulli(link = "logit"))
bf_mon  <- bf(monitoring ~ stress + importance + (1+stress+importance | id),
              family = bernoulli(link = "logit"))

# include random slope for offloading in PM
bf_pm   <- bf(prospective_memory ~ stress + importance + offloading + monitoring + 
                (1 + stress+importance+offloading+monitoring | id),
              family = bernoulli(link = "logit"))

mvform <- bf_off + bf_mon + bf_pm + set_rescor(FALSE) # no residual cpr for Bernoulli

# example priors (weakly informative)
priors <- c(
  prior(normal(0, 1), class = "b", resp = "offloading"),
  prior(normal(0, 1), class = "b", resp = "prospectivememory"),
  prior(normal(0, 1), class = "b", resp = "monitoring"),
  
  
  prior(normal(0, 1), class = "Intercept", resp = "prospectivememory")
)

# ---- Pilot fit: run one quick fit to inspect parameter names & check sampling ----
# (Use small chains for pilot; increase chains/iter for final sims)
 fit_pilot <- brm(mvform, data = dat_example, prior = priors,
                  chains = 2, iter = 1000, cores = 2,
                  control = list(adapt_delta = 0.95), refresh = 0)
 print(fit_pilot)
 pn<-names(as_draws_df(fit_pilot) ) # check exact parameter names: useful for detection step
 fixef(fit_pilot)

# ---- Detection function: example checks for offloading -> prospective_memory effect ----
# After running a pilot fit, check names from posterior_names() and adapt the pattern below.
 
 # typical brms name: "b_prospective_memory_offloading"
 PM_offload <- pn[grepl("b_prospectivememory_offloading", pn)]
 PM_monitoring <- pn[grepl("b_prospectivememory_monitoring", pn)]
 PM_stress <- pn[grepl("b_prospectivememory_stress", pn)]
 PM_importance <- pn[grepl("b_prospectivememory_importance", pn)]
 offloading_stress<- pn[grepl("b_offloading_stress", pn)]
 offloading_importance<- pn[grepl("b_offloading_importance", pn)]
 monitoring_importance<- pn[grepl("b_monitoring_importance", pn)]
 monitoring_stress<- pn[grepl("b_monitoring_stress", pn)]
 
 # the 8 patha
 vars<-c(PM_offload,PM_monitoring,PM_stress,PM_importance,offloading_stress,
         offloading_importance,monitoring_importance,monitoring_stress)
 
 
 df<- data.frame(matrix(ncol = length(vars), nrow = 0))
 colnames(df) <- vars
 
 
detect_effects <- function(fit, thresh = 0.95) {

 
 # the direction
 directions<-c(">", ">", "<", ">", "<",">", ">", "<")
 
  
  if (length(PM_offload) == 0) {
    stop("Could not automatically find parameter name for offloading -> pm.
         #Run pilot fit and inspect posterior_names(fit).")
  }
 
 # current posterior tests
 c_test<-vector()
 
 for (v in 1:length(vars)){
   
   
   # draw from postterior sampels
   c_draw<-as_draws_df(fit, variable = vars[v], fixed = TRUE)[[1]]
   
   # mean of the draws that pass the threshold
   c_dir<-directions[v]
   if (c_dir == ">") pass_thresh<-mean(c_draw >0) > thresh
   if (c_dir == "<") pass_thresh<-mean(c_draw <0) > thresh
   
   c_test[v]<-pass_thresh
   
 }
 v<-v+1
 
 # do the mediation effect
 path_a<-as_draws_df(fit, variable = "b_prospectivememory_offloading", fixed = TRUE)[[1]]
 path_b<-as_draws_df(fit, variable = "b_offloading_stress", fixed = TRUE)[[1]]
 
 offloading_stress_med<-path_a*path_b
 pass_thresh_off_stress<-mean(offloading_stress_med <0) > thresh
 
 c_test[v]<-pass_thresh_off_stress
 
 v<-v+1
 
 path_d<-as_draws_df(fit, variable = "b_offloading_importance", fixed = TRUE)[[1]]
 offloading_imp_med<-path_a*path_d
 pass_thresh_off_importance<-mean(offloading_imp_med >0) > thresh
 
 c_test[v]<-pass_thresh_off_importance
 
 v<-v+1
 
 path_e<-as_draws_df(fit, variable = "b_monitoring_stress", fixed = TRUE)[[1]]
 path_f<-as_draws_df(fit, variable = "b_prospectivememory_monitoring", fixed = TRUE)[[1]]
 
 monitor_stress_med<-path_e*path_f
 pass_thresh_mon_stress<-mean(monitor_stress_med <0) > thresh
 
 c_test[v]<-pass_thresh_mon_stress
 
 v<-v+1
 
 path_g<-as_draws_df(fit, variable = "b_monitoring_importance", fixed = TRUE)[[1]]
 monitor_import_med<-path_g*path_f
 pass_thresh_mon_import<-mean(monitor_import_med >0) > thresh
 
 c_test[v]<-pass_thresh_mon_import
 
 
return(c_test)
 
}
 

# ---- Power simulation wrapper  ----
run_power_sim_glmm <- function(initialN = 20, endN=220, step = 20, R = 10,
                               T_per = 14, chains = 1, iter = 100, seed = 1,
                               detect_fun = detect_effects) {
  # sample simulation
  sample_s<-seq(from = initialN,to =  endN, by = step)
  
  # create the dataset with the sample sizes
  dataset_power<- data.frame(matrix(ncol = (length(vars)+4)*2, nrow = length(sample_s)))
  
  all_vars<-c(vars, "mediation_stress_on_offloading", "mediation_imp_offloading", 
              "mediation_stress_on_monintoring", "mediation_imp_on_monitoring")
  
  names(dataset_power)<-c(all_vars, paste0("sd_",all_vars))
  
  dataset_power$sample_size<-sample_s
  
  set.seed(seed)
  detected <- vector()
  
  for (s in 1:length(sample_s)){
  
  for (r in seq_len(R)) {
    
    N<-sample_s[s]
    
    d <- simulate_ild_pm(N = N, T_per = T_per, seed = seed + r,
                           include_pm_off_slope = T)
    
    fit <- brm(mvform, data = d, prior = priors,
               chains = chains, iter = iter, cores = 1,
               control = list(adapt_delta = 0.95),
               refresh = 0)
    
    detected <- rbind(detected, detect_fun(fit))
    
    message(sprintf("[%d/%d] N=%d detected=%s", r, R, N,detect_fun(fit)))
  }
    
  # average across the columns
  power<-c(colMeans(detected), apply(detected,2, sd), sample_s[s])
  
  # append to the dataset
  dataset_power[s,]<-power
  }
  
  return(dataset_power)
  
}


power<-run_power_sim_glmm()

write.csv(power, file = "power.csv")

```

*The resulting power curves showed that power increased with sample size for all paths, but the mediator-to-outcome effects generally required larger samples than the direct effects and predictor-to-mediator effects.*

```{r}
#| eval: true
#| message: false
#| warning: false
#| include: true

library(dplyr)
library(tidyr)
library(ggplot2)

power_df <- read.csv("power.csv")

# delete the "X" variable
power_df<-power_df[!names(power_df) %in% "X",]

# delete the sd
sdvars<-grep("sd_", names(power_df))

power_df<-power_df[, -sdvars]

power_long <- power_df %>%
  pivot_longer(
    cols = -sample_size,
    names_to = "parameter",
    values_to = "power"
  ) %>%
  mutate( # rename parameters
    parameter = recode(
      parameter,
      b_prospectivememory_offloading = "PM ~ Offloading",
      b_prospectivememory_monitoring = "PM ~ Monitoring",
      b_prospectivememory_stress = "PM ~ Stress",
      b_prospectivememory_importance = "PM ~ Importance",
      b_offloading_stress = "Offloading ~ Stress",
      b_offloading_importance = "Offloading ~ Importance",
      b_monitoring_stress = "Monitoring ~ Stress",
      b_monitoring_importance = "Monitoring ~ Importance",
      mediation_stress_on_offloading= "stress~offloading~PM" ,
      mediation_imp_offloading=  "stress~offloading~PM" ,
      mediation_stress_on_monintoring = "stress~monitoring~PM",
     mediation_imp_on_monitoring =  "importance~monitoring~PM",
    )
  )

p <- ggplot(power_long, aes(x = sample_size, y = power, color = parameter)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_hline(yintercept = 0.80, linetype = "dashed") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    x = "Sample size",
    y = "Power",
    color = "Path",
    title = "Power analysis results"
  ) +
  theme_minimal(base_size = 13)

print(p)
ggsave("power_plot_labeled.png", p, width = 10, height = 6, dpi = 300)
```

## 
